
# Data Analysis

:::: {.panel-tabset .panel-tabset-pills}

```{r setup, include=FALSE}
library(tidyverse)
nps_data <- readRDS("nps_clean_data.rds")

most_visited_nps_species_data <- nps_data$species_data
density_df <- nps_data$density_data
```


## Threatened & Endangered Species Patterns
This section focuses on species that are at risk, such as those labeled as Endangered, Threatened, or Species of Concern. The examination of how many species exist in each national park, what proportion they represent out of the total species in that park, and how their nativeness varies was completed. These insights help identify parks with higher conservation burdens and explore which types of species are most vulnerable.

```{r}
library(tidyverse)

te_data <- most_visited_nps_species_data |>
  filter(TECategory != "Not Listed")

# TE species count by park
ggplot(te_data, aes(x = fct_infreq(ParkName), fill = TECategory)) +
  geom_bar(position = "dodge") +
  labs(title = "Threatened/Endangered Species by Park", x = "Park", y = "Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Percent TE species in each park
te_percent <- most_visited_nps_species_data |>
  group_by(ParkName) |>
  summarise(
    Total = n(),
    TE = sum(TECategory %in% c("Endangered", "Threatened", "Species of Concern"))
  ) |>
  mutate(PercentTE = TE / Total * 100)

ggplot(te_percent, aes(x = fct_reorder(ParkName, PercentTE), y = PercentTE)) +
  geom_col(fill = "firebrick") +
  labs(title = "Percentage of TE Species by Park", x = "Park", y = "% Threatened/Endangered") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# TE species by Nativeness
ggplot(te_data, aes(x = Nativeness, fill = TECategory)) +
  geom_bar(position = "fill") +
  labs(title = "TE Species by Nativeness", y = "Proportion") +
  scale_y_continuous(labels = scales::percent)

#  TE species by Category
ggplot(te_data, aes(x = fct_infreq(CategoryName))) +
  geom_bar(fill = "darkgreen") +
  labs(title = "TE Species by Category", x = "Category", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Nativeness and Ecological Integrity
Ecological integrity can be impacted by non-native species, which may disrupt the balance of natural ecosystems. This tab explores the percentage of non-native species in each park, the biological categories most commonly non-native, and their occurrence status. Understanding these patterns helps assess ecological health and informs management priorities.

```{r}
# Non-native species percent per park
nativeness_summary <- most_visited_nps_species_data |>
  group_by(ParkName) |>
  summarise(
    Total = n(),
    NonNative = sum(Nativeness == "Non-native", na.rm = TRUE)
  ) |>
  mutate(NonNative_Percent = NonNative / Total * 100)

ggplot(nativeness_summary, aes(x = fct_reorder(ParkName, NonNative_Percent), y = NonNative_Percent)) +
  geom_col(fill = "tomato") +
  labs(title = "Percent of Non-Native Species by Park", x = "Park", y = "% Non-Native") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Non-native by category
non_native_by_cat <- most_visited_nps_species_data |>
  filter(Nativeness == "Non-native")

ggplot(non_native_by_cat, aes(x = fct_infreq(CategoryName))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Non-Native Species by Category", x = "Category", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Non-native by Occurrence
ggplot(non_native_by_cat, aes(x = Occurrence)) +
  geom_bar(fill = "orange") +
  labs(title = "Occurrence Status of Non-Native Species", x = "Occurrence", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Species Richness and Biodiversity Composition
Biodiversity is a key indicator of ecological resilience. Here the overall species richness in each park which is the count of unique species observed was analyzed. Also, the taxonomic breakdown (e.g., mammals, birds, plants) was exploer  and highlighted how species composition varies across the parks. This helped reveal which parks are hotspots for biological diversity.

```{r}
# Total species richness per park
richness <- most_visited_nps_species_data |>
  group_by(ParkName) |>
  summarise(Species_Count = n_distinct(SciName))

ggplot(richness, aes(x = fct_reorder(ParkName, Species_Count), y = Species_Count)) +
  geom_col(fill = "darkorchid") +
  labs(title = "Species Richness by Park", x = "Park", y = "Unique Species") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Category breakdown by park
category_dist <- most_visited_nps_species_data |>
  group_by(ParkName, CategoryName) |>
  summarise(n = n(), .groups = "drop")

ggplot(category_dist, aes(x = ParkName, y = n, fill = CategoryName)) +
  geom_bar(stat = "identity") +
  labs(title = "Species Categories per Park", x = "Park", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Vertebrate vs Plant/Invertebrate
category_grouped <- most_visited_nps_species_data |>
  mutate(Group = case_when(
    CategoryName %in% c("Mammal", "Bird", "Reptile", "Amphibian", "Fish") ~ "Vertebrate",
    CategoryName == "Plant" ~ "Plant",
    TRUE ~ "Other"
  )) |>
  count(Group)

ggplot(category_grouped, aes(x = Group, y = n, fill = Group)) +
  geom_col() +
  labs(title = "Overall Taxonomic Composition", x = "Group", y = "Count")

# Mammal and Bird richness per park
mammal_bird_rich <- most_visited_nps_species_data |>
  filter(CategoryName %in% c("Mammal", "Bird")) |>
  group_by(ParkName, CategoryName) |>
  summarise(Species_Count = n_distinct(SciName), .groups = "drop")

ggplot(mammal_bird_rich, aes(x = ParkName, y = Species_Count, fill = CategoryName)) +
  geom_col(position = "dodge") +
  labs(title = "Mammal and Bird Richness by Park", x = "Park", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Park Area and Species Density
Do larger parks host more species? This section explores that question by visualizing the relationship between park area and species richness. We also calculate species density (species per km²) and map patterns across the U.S. These visualizations reveal which parks are biodiversity-dense and how park size influences ecological diversity.


```{r}
# Load US map data
us_map <- map_data("state")

# US map - Species Count and Density
ggplot() +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group), 
               fill = "gray90", color = "white") +
  geom_point(data = density_df, aes(x = Longitude, y = Latitude, 
                                    size = Species_Density, 
                                    color = Species_Count)) +
  geom_text(data = density_df, aes(x = Longitude, y = Latitude, label = ParkName),
            hjust = -0.1, size = 3) +
  scale_color_gradient(low = "orange", high = "purple") +
  scale_size(range = c(3, 10)) +
  labs(
    title = "Species Richness and Density Across National Parks",
    x = "Longitude", y = "Latitude",
    color = "Species Count", size = "Species/km²"
  ) +
  theme_minimal()

# US map - Park Area as size, Density as color
ggplot() +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group), 
               fill = "gray95", color = "white") +
  geom_point(data = density_df, aes(x = Longitude, y = Latitude, 
                                    size = ParkArea_km2, 
                                    color = Species_Density)) +
  geom_text(data = density_df, aes(x = Longitude, y = Latitude, label = ParkName),
            hjust = -0.1, size = 3) +
  scale_color_viridis_c() +
  scale_size(range = c(3, 10)) +
  labs(
    title = "National Park Area and Species Density",
    x = "Longitude", y = "Latitude",
    color = "Density (species/km²)", size = "Area (km²)"
  ) +
  theme_minimal()

```
::::